Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-22 paper/ms.tex
OMG GET RID OF ALL THE 'will'S

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that scale well with amount of data.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the application of this method by applying it to simulated and
real astronomical time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.  The power spectrum of an irregularly sampled time series may be
estimated with Bayesian inference, with a posterior estimate of the associated
uncertainties, in principle avoiding the problems associated with window functions
and aliasing.
The method is flexible, fast, and most importantly, interpretable.
There are a wide range of potential applications within astronomical data
analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic
nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987,Bond:1999}, correlated instrumental noise
\citep{Gibson:2012}, spectroscopic calibration \citep{Czekala:2017,Evans:2015}
and residuals caused by model inconsistencies (CITE + better words).
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
$\sim10^{3-5}$ observations each~---~this cost is prohibitive.

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).
This method is especially appealing compared to other similar methods~--~we
will return to these below~--~because it is exact, flexible, simple, and fast.

In the following pages, we will motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released well tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

@@

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the computational
cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
is $\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
This means that for large datasets, every evaluation of the likelihood will
quickly become computationally intractable.
In this case, standard non-linear optimization or MCMC will no longer be
practical inference methods.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse will be dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
in the mixture and $N$ is still the number of data points.

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
$c_j \to c_j\pm i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(N)$ operations.
The details of this method and a few implementation considerations are
discussed in the following \sectionname\ but we will first discuss some
properties of this covariance function.

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we will refer to the argument within the sum as a ``\celerite\
term'' for the remainder of this paper.
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

@@

To provide some insight for this method, we will follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and this is the model that was studied by
\citet{Rybicki:1995}.
To demonstrate this method, we will write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we will follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ that we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

In the following section, we will discuss this method more generally but it's
worth noting a few important facts that can already be seen in this example.
First, the fundamental reason why this matrix $K$ can be solved efficiently is
the following property of exponentials
\begin{eqnarray}
    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
    e^{-c\,(t_3 - t_1)}
\end{eqnarray}
and it is important to note that this property does not extend to other common
covariance functions like the ``exponential-squared'' function
\begin{eqnarray}
    k(\tau) \propto e^{-c\,\tau^2} \quad.
\end{eqnarray}
Furthermore, our derivation of the extended matrix requires that the data
points be monotonically sorted in time.
Neither of these properties will be satisfied in general for multidimensional
inputs and all of our following discussion will assume a sorted
one-dimensional dataset.

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix will still be banded with a bandwidth that scales linearly
with the number of components $J$.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
This means that we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(N)$ operations.

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we will take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celerite\ method
described in the previous section.

@@

To demonstrate the application of \celerite, we will start by inferring
posterior constraints on the parameters of a GP model applied to several
simulated datasets with known properties.
In the following section, we will expand on these examples by applying
\celerite\ to real datasets.
In the first example, we demonstrate that \celerite\ can be used to measure
the power spectrum of a process when the data are generated from a \celerite\
model.
In the second example, we demonstrate that \celerite\ can be used as an
effective model even if the true process cannot be represented in the space of
allowed models.
This is an interesting example because, when analyzing real data, we rarely
have any fundamental reason to believe that the data were generated by a GP
model with a specific kernel.
Even in these cases, GPs can be useful effective models and \celerite\
provides computational advantages over other GP methods.

@@

In this first example, we will simulate a dataset using a known \celerite\
process and fit it with \celerite\ to demonstrate that valid inferences can be
made in this idealized case.
The simulated dataset is shown in the left panel of \Figure{simulated-correct}
and it was generated using a SHO kernel (\eqalt{sho-kernel}) with parameters
$S_0 = 1$, $\omega_0 = e^2$, and $Q = e^2$.
The true PSD is shown as a dashed line in the right panel of
\Figure{simulated-correct}.
We applied log-uniform priors to all of the parameters and used \emcee\
\citep{Foreman-Mackey:2013} to sample the joint posterior density and computed
the marginalized posterior inference of the PSD.
This inference is shown in the right panel of \Figure{simulated-correct} as a
blue contour indicating 68\% of the posterior mass.
It is clear from this figure that, as expected, the inference correctly
reproduces the true PSD.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celerite\ process.
This means that the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we will use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.

>>>>

The growing field of large-scale time domain astronomy requires methods for
probabilistic inference that are computationally tractable even with large
datasets.
We present a method for Gaussian Process regression in one dimension with a
computational cost that scales linearly with the size of the dataset.
This method exploits structure in the problem when the covariance function is
expressed as a mixture of complex exponentials, without requiring evenly
spaced observations or uniform noise.
This form of covariance arises naturally when the process is a mixture of
stochastically-driven damped harmonic oscillators~--~providing a physical
motivation for and interpretation of this choice~--~but we also demonstrate
that it is effective in many other cases.
We present a mathematical description of the method, the details of the
implementation, and a comparison to existing scalable Gaussian Process methods.
We demonstrate the method by applying it to simulated and real astronomical
time series datasets.
These demonstrations are examples of probabilistic inference of stellar
rotation periods, asteroseismic oscillation spectra, and transiting planet
parameters.
In all of these cases, we infer posterior distributions over models for the
power spectrum of the process that generated these irregularly sampled time
series without computing a standard algebraic estimator.
The method is flexible, fast, and most importantly, interpretable, with a wide
range of potential applications within astronomical data analysis and beyond.
We provide well tested and documented open-source implementations of this
method in \project{C++}, \project{Python}, and \project{Julia}.

@@

Gaussian Processes \citep[GPs;][]{Rasmussen:2006} are popular stochastic
models for time-series analysis.
For GP modeling, a functional form is chosen to describe the autocovariance
of the data and the parameters of this function are fit for or marginalized.
In the astrophysical literature, GPs have been used to model stochastic
variability in light curves of stars \citep{Brewer:2009}, active galactic
nuclei \citep{Kelly:2014}, and the logarithmic flux of X-ray binaries
\citep{Uttley:2005}.
They have also been used as models for the cosmic microwave background
\citep{Bond:1987,Bond:1999}, correlated instrumental noise
\citep{Gibson:2012}, and spectroscopic calibration
\citep{Czekala:2017,Evans:2015}.
While these models are widely applicable, their use has been limited, in
practice, by the computational cost and scaling.
In general, the cost of computing a GP likelihood scales as the third power of
the number of data points $\mathcal{O}(N^3)$ and in the current era of large
time domain surveys~--~with as many as $\sim10^{4-9}$ targets with
$\sim10^{3-5}$ observations each~---~this cost is prohibitive.

In this paper, we present a method for computing a class of GP models that
scales linearly with the number of data points $\mathcal{O}(N)$ for one
dimensional data sets.
This method is a generalization of a method developed by
\citet{Ambikasaran:2015} that was, in turn, built on intuition from a twenty
year old paper \citep{Rybicki:1995}.
For this method to be applicable, the data must be one-dimensional and the
covariance function must have a specific form.
However, there is no further constraint on the data or the model.
In particular, the measurements don't need to be evenly spaced and the
uncertainties can be heteroscedastic (non-uniform).
This method is especially appealing compared to other similar methods~--~we
return to these below~--~because it is exact, flexible, simple, and fast.

In the following pages, we motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released well tested and documented open source
implementations written in \project{C++}, \project{Python}, and
\project{Julia}.
These are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

@@

GP models have been widely used across the physical sciences but their
application is generally limited to small datasets because the computational
cost of computing the inverse and determinant of the matrix $K_\bvec{\alpha}$
is $\mathcal{O}(N^3)$.
In other words, this cost is proportional to the cube of the number of data
points $N$.
This means that for large datasets, every evaluation of the likelihood
quickly becomes computationally intractable.
In this case, standard non-linear optimization or MCMC will no longer be
practical inference methods.

@@

To scale GP models to larger datasets, \citet{Rybicki:1995} presented a method
of computing the first term in \eq{gp-likelihood} in $\mathcal{O}(N)$
operations when the covariance function is given by
\begin{eqnarray}\eqlabel{kernel-simple}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})
\end{eqnarray}
where $\{{\sigma_n}^2\}_{n=1}^N$ are the measurement uncertainties,
$\delta_{nm}$ is the Kronecker delta, and $\bvec{\alpha} = (a,\,c)$.
The intuition behind this method is that, for this choice of $k_\bvec{\alpha}$,
the inverse of $K_\bvec{\alpha}$ is tridiagonal and can be computed
with a small number of operations for each data point.
Subsequently, \citet{Ambikasaran:2015} generalized this method to arbitrary
mixtures of exponentials
\begin{eqnarray}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J a_j\,\exp(-c_j\,\tau_{nm})\quad.
\end{eqnarray}
In this case, the inverse is dense but \eq{gp-likelihood} can still be
evaluated in $\mathcal{O}(N)$ operations where $J$ is the number of components
in the mixture and $N$ is still the number of data points.

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j\pm i\,b_j$ and
$c_j \to c_j\pm i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(N)$ operations.
The details of this method and a few implementation considerations are
discussed in the following \sectionname\ but we first discuss some
properties of this covariance function.

By rewriting the exponentials in \eq{celerite-kernel-complex} as sums of sine
and cosine functions, we can see the autocorrelation structure is defined by a
mixture of quasiperiodic oscillators
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
For clarity, we refer to the argument within the sum as a ``\celerite\
term'' for the remainder of this paper.
The Fourier transform\footnote{Here and throughout we have defined the Fourier
transform of the function $f(t)$ as $F(\omega)={(2\,\pi)}^{-1/2}\,
\int_{-\infty}^\infty f(t)\,e^{i\,\omega\,t}\dd t$.} of this covariance
function is the power spectral density (PSD) of the process and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+{({c_j}^2+{d_j}^2)}^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
return to a more general discussion of the physical intuition in a moment
but we start with a discussion of some useful special cases of this model.

@@

To provide some insight for this method, we follow
\citet{Ambikasaran:2015} and start by working through a simple example.
In this case, we'll assume that we have three data points
$\{y_1,\,y_2,\,y_3\}$ observed at times $\{t_1,\,t_2,\,t_3\}$ with measurement
variances $\{{\sigma_1}^2,\,{\sigma_2}^2,\,{\sigma_3}^2\}$ and we would
like to compute the likelihood of these data under a GP model with the
covariance function
\begin{eqnarray}
k(\tau_{nm}) = \sigma_n^2\,\delta_{nm} + a\,\exp(-c\,\tau_{nm})\quad.
\end{eqnarray}
This is the result of setting $J=1$, $b=0$, and $d=0$ in
\eq{celerite-kernel-complex} and this is the model that was studied by
\citet{Rybicki:1995}.
To demonstrate this method, we write out the full system of
equations that we must solve to apply the inverse of $K$ in order to compute
the first term of \eq{gp-likelihood}.
In matrix notation, this can be written as
\begin{eqnarray}\eqlabel{impl-matrix}
K\,\bvec{z} &=& \bvec{y}, \\
\begin{pmatrix}
    a+{\sigma_1}^2 & a\,e^{-c\,\tau_{2,1}} & a\,e^{-c\,\tau_{3,1}}\\
    a\,e^{-c\,\tau_{2,1}} & a+{\sigma_2}^2 & a\,e^{-c\,\tau_{3,2}}\\
    a\,e^{-c\,\tau_{3,1}} & a\,e^{-c\,\tau_{3,2}} & a+{\sigma_3}^2
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ z_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ y_2 \\ y_3
\end{pmatrix}
\end{eqnarray}
where our goal is to solve for the unknown vector \bvec{z} for a given matrix
$K$ and vector \bvec{y}.
In this equation, we have assumed that the mean function is zero but a
non-zero mean could be included by replacing \bvec{y} by
$\bvec{r}_\bvec{\theta}$ as defined in \sect{gps}.
Now, if we introduce the variables
\begin{eqnarray}\eqlabel{algo-first}
    u_n = e^{-c\,\tau_{n+2,n+1}}\,u_{n+1} + a\,z_{n+1}
\end{eqnarray}
where $u_{N} = 0$, and
\begin{eqnarray}
    g_n = e^{-c\,\tau_{n+1,n}}\,g_{n-1} + e^{-c\,\tau_{n+1,n}}\,z_{n}
\end{eqnarray}
where $g_{0} = 0$,
the system of equations can be rewritten as
\begin{eqnarray}
(a+{\sigma_1}^2)\,z_1 + e^{-c\,\tau_{2,1}}\,u_1 &=& y_1 \\
a\,g_1 + (a+{\sigma_2}^2)\,z_2 + e^{-c\,\tau_{3,2}}\,u_2 &=& y_2 \\
a\,g_2 + (a+{\sigma_3}^2)\,z_3 &=& y_3 \quad. \eqlabel{algo-last}
\end{eqnarray}
The system defined by \eq{algo-first} through \eq{algo-last} can be
rewritten as a matrix equation to show the benefit that this seemingly trivial
reformulation provides:
\begin{eqnarray}
\begin{pmatrix}
    a+{\sigma_1}^2 & e^{-c\,\tau_{2,1}} & 0 & 0 & 0 & 0 & 0 \\
    e^{-c\,\tau_{2,1}} & 0 & -1 & 0 & 0 & 0 & 0 \\
    0 & -1 & 0 & a & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & a & a+{\sigma_2}^2 & e^{-c\,\tau_{3,2}} & 0 & 0 \\
    0 & 0 & e^{-c\,\tau_{3,2}} & e^{-c\,\tau_{3,2}} & 0 & -1 & 0 \\
    0 & 0 & 0 & 0 & -1 & 0 & a \\
    0 & 0 & 0 & 0 & 0 & a & a+{\sigma_3}^2 \\
\end{pmatrix}\,
\begin{pmatrix}
    z_1 \\ u_1 \\ g_1 \\ z_2 \\ u_2 \\ g_2 \\ z_3
\end{pmatrix} &=&
\begin{pmatrix}
    y_1 \\ 0 \\ 0 \\ y_2 \\ 0 \\ 0 \\ y_3
\end{pmatrix}\nonumber
\end{eqnarray}
where we follow \citet{Ambikasaran:2015} and call this the ``extended''
system and rewrite \eq{impl-matrix} as
\begin{eqnarray}
    K_\mathrm{ext}\,\bvec{z}_\mathrm{ext} &=& \bvec{y}_\mathrm{ext} \quad.
\end{eqnarray}
Even though $K_\mathrm{ext}$ is a larger matrix than the $K$ that we started
with, it is now sparse with banded structure that can be exploited to solve
the system efficiently.
In particular, sparse solvers are available that can perform a
LU-decomposition of matrices like this in $\mathcal{O}(N)$
operations~--~instead of the $\mathcal{O}(N^3)$ that would be required in
general~--~and we can take advantage of these algorithms to solve our system
exactly because the target vector $\bvec{z}$ is a subset of the elements of
$\bvec{z}_\mathrm{ext}$.

In the following section, we discuss this method more generally but it's
worth noting a few important facts that can already be seen in this example.
First, the fundamental reason why this matrix $K$ can be solved efficiently is
the following property of exponentials
\begin{eqnarray}
    e^{-c\,(t_3 - t_2)} \, e^{-c\,(t_2 - t_1)} =
    e^{-c\,(t_3 - t_2 + t_2 - t_1)} =
    e^{-c\,(t_3 - t_1)}
\end{eqnarray}
and it is important to note that this property does not extend to other common
covariance functions like the ``exponential-squared'' function
\begin{eqnarray}
    k(\tau) \propto e^{-c\,\tau^2} \quad.
\end{eqnarray}
Furthermore, our derivation of the extended matrix requires that the data
points be monotonically sorted in time.
Neither of these properties be satisfied in general for multidimensional
inputs and all of our following discussion assumes a sorted
one-dimensional dataset.

\citet{Ambikasaran:2015} demonstrated two key facts that allow us to use this
extended matrix formalism in practice.
First, even if the covariance function is a mixture of exponentials, the
extended matrix is still be banded with a bandwidth that scales linearly
with the number of components $J$.
Second, \citet{Ambikasaran:2015} proved that the determinant of
$K_\mathrm{ext}$ is equal to the determinant of $K$ up to a sign.
This means that we can use this extended matrix formalism to compute the
marginalized likelihood in $\mathcal{O}(N)$ operations.

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celerite\ method
described in the previous section.

@@

To demonstrate the application of \celerite, we start by inferring
posterior constraints on the parameters of a GP model applied to several
simulated datasets with known properties.
In the following section, we expand on these examples by applying
\celerite\ to real datasets.
In the first example, we demonstrate that \celerite\ can be used to measure
the power spectrum of a process when the data are generated from a \celerite\
model.
In the second example, we demonstrate that \celerite\ can be used as an
effective model even if the true process cannot be represented in the space of
allowed models.
This is an interesting example because, when analyzing real data, we rarely
have any fundamental reason to believe that the data were generated by a GP
model with a specific kernel.
Even in these cases, GPs can be useful effective models and \celerite\
provides computational advantages over other GP methods.

@@

In this first example, we simulate a dataset using a known \celerite\
process and fit it with \celerite\ to demonstrate that valid inferences can be
made in this idealized case.
The simulated dataset is shown in the left panel of \Figure{simulated-correct}
and it was generated using a SHO kernel (\eqalt{sho-kernel}) with parameters
$S_0 = 1$, $\omega_0 = e^2$, and $Q = e^2$.
The true PSD is shown as a dashed line in the right panel of
\Figure{simulated-correct}.
We applied log-uniform priors to all of the parameters and used \emcee\
\citep{Foreman-Mackey:2013} to sample the joint posterior density and computed
the marginalized posterior inference of the PSD.
This inference is shown in the right panel of \Figure{simulated-correct} as a
blue contour indicating 68\% of the posterior mass.
It is clear from this figure that, as expected, the inference correctly
reproduces the true PSD.

@@

For this example, we simulate a dataset using a known GP model with a kernel
outside of the support of a \celerite\ process.
This means that the true autocorrelation of the process can never be correctly
represented by the model that we're using to fit but we use this example
to demonstrate that, at least in this case, valid inferences can still be made
about the physical parameters of the model.
